---
title: "Workforce Cost Curves: Acute care programme"
execute: 
  echo: false
format: 
   html:
    embed-resources: true
    df-print: paged
    toc: true
---

[For Management Information only, not for onward distribution.]{style="color:red;"}

```{r}
#| output: false

# set-up

# data retrieval and wrangling tools
library(haven)
library(rlist)
library(readr)
library(janitor)
library(dplyr)
library(tidyr)
library(odbc)

# data viz tools for pretty interactive graphs
library(ggplot2)
library(patchwork)
library(ggiraph)
library(viridis)
library(scales)

# load custom functions
for (filename in list.files(
  "functions", pattern = "\\.R", full.names = TRUE, 
  recursive = TRUE)
) {
  source(filename)
}

# load reference files
cbdcs_column_dictionary_sfr_53 <- read_csv("reference_files/cbdcs_column_dictionary.csv", show_col_types = FALSE) |> 
  clean_names() |> 
  filter(cd_formname == "SFR 5.3") |> 
  remove_constant() |> 
  mutate(cd_snakecase_description = make_clean_names(cd_description), 
         cd_columnno = as.character(paste0("col", cd_columnno)))

#if outputs directory doesnt exist create it
if(!(dir.exists("./outputs"))){
  dir.create("./outputs")
}

# import the fixed and variable cost proportions for the Acute care programme
fixed_variable_prop <- read_sav("data/Fixed_Variable_Ratios_Acute.sav") |> 
  # mutate(across(where(is.numeric), ~.x/100)) |> 
  rename("fixed_costs_prop" = "fixed", "variable_costs_prop" = "variable")

# total expenditure cost per head
total_exp_cost_per_head <- readRDS("data/acute_cost_per_head_v26_27.rds") |>
  mutate(acute_cph = round(acute_cph, 2), sex = as.integer(sex)) |> 
  add_age_labels() |> 
  add_sex_label()
```

## Aim

To calculate the workforce cost per head by following the same methodology as used in the NRAC formula to derive the total cost per head. For clarity, this refers to the cost per capita, not the cost per patient.

## Caveats Summary

-   Workforce costs are derived using the same fixed/variable cost proportions that are used to calculate total expenditure cost per head in the NRAC formula. These are applied to inpatient data only and determine the proportion of cost attributable to discharges (fixed) and bed days (variable).
-   The fixed/variable proportions are intended to act as case-mix adjustment but are limited to the extent that they do not precisely attach greater cost to episodes of care which are more resource intensive. This should be given consideration when interpreting the workforce or total costs per head presented or using them in other analysis.
-   NHS Borders costs are excluded from these data since they were unable to submit a full SFR 5.3 dataset for financial year 23/24 (see [23/24 Scottish Health Service Costs](https://publichealthscotland.scot/media/34037/2025-02-25-costs-report_final.pdf) p.21).
-   Combines workforce costs from SFRs: 5.3, 5.5, 5.7, 5.7N and 5.9. Apart from Medical, Dental and Nursing, all other staff costs include supplies.
-   Uses activity data A&E, SMR00 (outpatients), and SMR01 (inpatients and daycases) to apportion costs.
-   Cost per head is derived by dividing the total cost for each age/sex group by the 2024 mid-year population estimates provided by NRS.

## 23/24 Acute Workforce Costs Data

Workforce costs by line number were derived following the same methodology as for the total expenditure used in the NRAC formula.

These workforce costs data are sourced from various Scottish Financial Returns (SFRs), see table below:

| Form Name | Description | SFR columns (staff type costs) |
|----|----|----|
| SFR 5.3 | Inpatient costs | 7-12 |
| SFR 5.5 | Day Case costs | 5-10 |
| SFR 5.7 | Outpatients treated in consultant-led clinics | 6-11 |
| SFR 5.7N | Outpatients treated in nurse-led clinics | 4-7 |
| SFR 5.9 | Day Patients | 4-8 |

***NHS Borders costs are excluded from these data since they were unable to submit a full SFR dataset for financial year 23/24 (see [23/24 Scottish Health Service Costs](https://publichealthscotland.scot/media/34037/2025-02-25-costs-report_final.pdf) p.21)***

```{r}
id_df <- read_csv(
  "/PHI_conf/Finance/CostsBook/Reference Files/userid.csv",
  col_types = "c"
)
uid <- id_df$username
pwd <- id_df$password

id_colnames <- c("formname", "provider", "location", "lineno")

# get costs data
sfr_53 <- get_costs_data(2023, "SFR 5.3") |>
  select(all_of(id_colnames), col3, c(col7:col12)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

# rename columns to something humans can understand
colnames(sfr_53) <- human_cbdcs_colnames(sfr_53) 

sfr_55 <- get_costs_data(2023, "SFR 5.5")|> 
  select(all_of(id_colnames), col1, c(col5:col10)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_55) <- human_cbdcs_colnames(sfr_55)
# sfr_55

sfr_57 <- get_costs_data(2023, "SFR 5.7")|> 
  select(all_of(id_colnames), col1, c(col6:col11)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_57) <- human_cbdcs_colnames(sfr_57)
# sfr_57

sfr_57n <- get_costs_data(2023, "SFR 5.7N")|> 
  select(all_of(id_colnames), col1, c(col4:col7)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_57n) <- human_cbdcs_colnames(sfr_57n)
# sfr_57n

sfr_59 <- get_costs_data(2023, "SFR 5.9")|> 
  select(all_of(id_colnames), col1, c(col4:col8)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_59) <- human_cbdcs_colnames(sfr_59)
# sfr_59


```

Here's an example of what the SFR data looks like:

```{r}
sfr_53
```

The SFR line numbers are recoded to map these costs to the correct specialties in the Scottish Morbidity Records (SMR) activity data. Then, workforce costs are aggregated at specialty level (i.e. aggregated by SFR line number) and patient type (inpatient, day case or outpatient). We have limited data on which proportion of the workforce costs reported are fixed (e.g. core staff), and which are variable (e.g. NHS staff bank, overtime, etc.). Therefore, we assume that the proportions of fixed/variable costs per Cost Book line number are the same as for total expenditure costs.

Here is an overview of the aggregated workforce costs by patient type, line number and staff type:

```{r}
sfr_list <- list(sfr_53, sfr_55, sfr_57, sfr_57n, sfr_59)

output_list <- lapply(sfr_list, parse_cost_by_lineno)

workforce_costs_by_lineno <- do.call(rbind, output_list) |> 
  
  # sum costs and activity by ipdc, lineno and staff type
  group_by(ipdc, lineno, staff_type) |> 
  summarise(across(where(is.numeric), ~sum(.)), .groups = "drop") |> 
  
  # join fixed and variable cost proportions
  left_join(fixed_variable_prop, by = c("ipdc", "lineno")) |> 
  
  # calculate fixed and variable costs
  mutate(fixed_cost = cost * fixed_costs_prop / 100, 
         variable_cost = cost * variable_costs_prop / 100)

workforce_costs_by_lineno

saveRDS(workforce_costs_by_lineno, "outputs/workforce_costs_by_lineno.rds")
```

## Workforce cost per head estimates

The aggregated workforce costs data is joined with the A&E, Outpatients (SMR00) and Inpatients & Day Cases (SMR01) activity data. A total cost is apportioned to each age/sex group based on the activity they account for and the fixed (number of discharges) and variable (number of bed days) breakdown. Each cost is then divided by the number of people in each age/sex group in the mid-year population estimates, giving an estimated cost per head.

This is executed using the same methodology as the NRAC formula for total expenditure cost per head.

```{r}

workforce_cost_per_head <- readRDS("data/acute_cost_per_head_v26_27_workforce.rds") |> 
  mutate(acute_cph = round(acute_cph, 1)) |> 
  add_staff_label() |> 
  add_age_labels() |> 
  add_sex_label() |> 
  mutate(sex = as.integer(sex))

# save a csv copy of acute workforce cost per head for SG
write_csv(workforce_cost_per_head, "outputs/acute_cost_per_head_v26_27_workforce.csv")

# overview of the cost per head by staff type age group and sex
workforce_cost_per_head |> select(staff_type_label, agegrp_label, sex_label, acute_cph)

```

## Plots

### Acute Costs Curves by Staff Type

```{r}

staff_types <- unique(workforce_cost_per_head$staff_type_label)

# create a cost-curves plot for each staff type
acute_plots <- list()

for(i in 1:length(staff_types)){
  
  staff_type_i <- staff_types[i]
  
  data_i <- workforce_cost_per_head |> 
                filter(staff_type_label == staff_type_i)
  
  plot_i <- plot_cost_curves(x = data_i, x_title = staff_type_i)
  
  # add plot to list  
  acute_plots <- list.append(acute_plots, plot_i)
  
  names(acute_plots)[i] <- staff_type_i
}

```

Costs curves for the Acute care programme across all staff types.

```{r}

girafe_plot_options(acute_plots[[which(names(acute_plots) == "All Staff")]])
```

Breakdown of the cost per head by staff type.

```{r}
# combine all the plots except for all staff
patchwork <- wrap_plots(acute_plots[which(names(acute_plots) != "All Staff")], 
                        ncol = 2,
                        nrow = 3,
                        axis_titles = "collect", 
                        guides = "collect")

girafe_plot_options(patchwork)

```

### Workforce Expenditure vs Total Expenditure

Does the acute cost per head for workforce costs follow the same trend as the cost per head for total expenditure?

```{r}

# create a dataframe with workforce data for all staff and data for total expenditure
comparison_df <- workforce_cost_per_head |> 
  filter(staff_type == "all_staff") |> 
  select(-staff_type, -staff_type_label) |> 
  left_join(total_exp_cost_per_head, by = c("agegrp", "sex", "agegrp_label", "sex_label"), suffix = c("_workforce", "_total")) |> 
  pivot_longer(c("acute_cph_workforce", "acute_cph_total"), names_to = "costs_type", values_to = "acute_cph") |> 
  mutate(costs_type = if_else(costs_type == "acute_cph_workforce", "workforce", "total expenditure"))
  
# filter females
comparison_df_f <- comparison_df |> 
  filter(sex == "2")

# filter males
comparison_df_m <- comparison_df |> 
  filter(sex == "1")


# plot one chart for females and one chart for males
p_f <- plot_total_vs_workforce(comparison_df_f, "Acute cost per head for females, total expenditure VS workforce" )
p_m <- plot_total_vs_workforce(comparison_df_m, "Acute cost per head for males, total expenditure VS workforce" )

girafe_plot_options(p_f)
girafe_plot_options(p_m)

```

### Proportion of the total cost per head which is attributed to workforce costs

What is the percentage of the total cost per head that can be attributed to staff costs? The charts below explore how the share of staff costs evolves across different staff types, age groups and sexes.

```{r}
proportion_df <- workforce_cost_per_head |> 
  left_join(total_exp_cost_per_head, by = c("agegrp", "sex", "agegrp_label", "sex_label"), suffix = c("_workforce", "_total")) |> 
  # mutate(prop_of_total_exp = round(acute_cph_workforce/acute_cph_total, 2))
  mutate(prop_of_total_exp = acute_cph_workforce/acute_cph_total)

p_f_prop <- plot_prop_of_total_expenditure(proportion_df |> filter(sex == 2), 
                                      "Proportion of total cost per head by age and staff type for females")
p_m_prop <- plot_prop_of_total_expenditure(proportion_df |> filter(sex == 1), 
                                      "Proportion of total cost per head by age and staff type for males")


girafe_plot_options(p_f_prop)
girafe_plot_options(p_m_prop)

```
