---
title: "Workforce Cost Curves: Acute care programme"
format: 
   html:
     df-print: paged
     toc: true
---

**Main Goal: Plot the cost per head by age band and sex, for the NRAC acute care programme. These outputs rely on the NRAC cost per head calculations**

Throughout this report we split this main goal into smaller goals to make things easier to digest.

## Set-up

```{r}
#| output: false

# set-up
library(readr)
library(janitor)
library(dplyr)
library(tidyr)
library(odbc)
library(ggplot2)

# load functions
for (filename in list.files(
  "functions", pattern = "\\.R", full.names = TRUE, 
  recursive = TRUE)
) {
  source(filename)
}

# load reference files
cbdcs_column_dictionary_sfr_53 <- read_csv("reference_files/cbdcs_column_dictionary.csv") |> 
  clean_names() |> 
  filter(cd_formname == "SFR 5.3") |> 
  remove_constant() |> 
  mutate(cd_snakecase_description = make_clean_names(cd_description), 
         cd_columnno = as.character(paste0("col", cd_columnno)))

```

## Reproduce the NRAC 26/27 Acute costs curve for all Acute costs

Load the acute cost per head output from the NRAC 26/27 publication. Then, plot the cost per head by age group and sex.

The acute cost per head output is sourced from <https://github.com/Public-Health-Scotland/NRAC/blob/develop/R/2-HCHS/1-age-sex/2-cost-weights/1-acute_cost_weights.R>.

```{r}
file_paths <- list.files("data", full.names = TRUE)
names(file_paths) <- list.files("data")

cost_per_head <- lapply(file_paths, readRDS)

ggplot(cost_per_head[[2]], aes(x = agegrp, y = acute_cph, color = sex))+
  geom_line()
```

Let's attempt to produce a similar output for workforce costs only.

## 23/24 Acute Workforce Costs Data

Goal: The goal in this section is to calculate total workforce costs by line number, following a similar methodology as in the [NRAC hospital activity costs](https://github.com/Public-Health-Scotland/NRAC/blob/develop/R/2-HCHS/1-age-sex/1-costs/3-hospital_activity_costs.R)

We source the necessary workforce costs data from various Scottish Financial Returns (SFR), see table below:

| Form Name | Description | Columns |
|-------------------|-------------------|----------------------------------|
| SFR 5.3 | Inpatient costs | col 3 (number of discharges), col 7-13 (costs by staff type) |
| SFR 5.5 | Day Case costs | col 1 (number of attendances), col 5-10 (costs by staff type) |
| SFR 5.7 | Outpatients treated in consultant clinics | col 1 (number of attendances), col 6-11 (costs by staff type) |
| SFR 5.7N | Outpatients treated at nurse-led clinics | col 1 (number of attendances), col 4-7 (costs by staff type) |
| SFR 5.9 | Day Patients | col 1 (number of attendances), col 4-8 (costs by staff type) |

***NHS Borders costs are excluded from these data since they were unable to submit a full SFR dataset for financial year 23/24 (see [23/24 Scottish Health Service Costs](https://publichealthscotland.scot/media/34037/2025-02-25-costs-report_final.pdf) p.21)***

```{r}
id_df <- read_csv(
  "/PHI_conf/Finance/CostsBook/Reference Files/userid.csv",
  col_types = "c"
)
uid <- id_df$username
pwd <- id_df$password

id_colnames <- c("formname", "provider", "location", "lineno")

# get costs data
sfr_53 <- get_costs_data(2023, "SFR 5.3") |>
  select(all_of(id_colnames), col3, c(col7:col12)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

# rename columns to something humans can understand
colnames(sfr_53) <- human_cbdcs_colnames(sfr_53) 
sfr_53

sfr_55 <- get_costs_data(2023, "SFR 5.5")|> 
  select(all_of(id_colnames), col1, c(col5:col10)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_55) <- human_cbdcs_colnames(sfr_55)
# sfr_55

sfr_57 <- get_costs_data(2023, "SFR 5.7")|> 
  select(all_of(id_colnames), col1, c(col6:col11)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_57) <- human_cbdcs_colnames(sfr_57)
# sfr_57

sfr_57n <- get_costs_data(2023, "SFR 5.7N")|> 
  select(all_of(id_colnames), col1, c(col4:col7)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_57n) <- human_cbdcs_colnames(sfr_57n)
# sfr_57n

sfr_59 <- get_costs_data(2023, "SFR 5.9")|> 
  select(all_of(id_colnames), col1, c(col4:col8)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_59) <- human_cbdcs_colnames(sfr_59)
# sfr_59


```

Next, let's recode SFR line numbers to map these costs to the correct specialties in the Scottish Morbidity Records(SMR) activity data. Finally, let's aggregate the workforce costs at specialty level (ie. aggregate costs by line number and patient type). Here we assume that the proportion of the costs that are fixed is 1. We assume this because we have limited data on which proportion of the workforce costs reported are fixed (eg. core staff), and which are variable (eg. NHS staff bank, overtime, etc.). This is all neatly packaged in the function `parse_cost_by_lineno` and iterated over all the SFRs.

```{r}
sfr_list <- list(sfr_53, sfr_55, sfr_57, sfr_57n, sfr_59)

output_list <- lapply(sfr_list, parse_cost_by_lineno)

workforce_costs_by_lineno <- do.call(rbind, output_list)
workforce_costs_by_lineno

saveRDS(workforce_costs_by_lineno, "outputs/workforce_costs_by_lineno.rds")
```

## Calculate cost per head

Now that the data is aggregated by patient type and line number, we can more accurately match the costs for the financial year to A&E, Outpatients (SMR00) and Inpatients & Day Cases (SMR01) activity data. The activity data is used to derive a total cost per age-sex group, then these costs are divided by the mid year population estimates, giving an estimated cost per head.

The above is executed by a modified version of the NRAC Acute costs weights script which I adapted to read-in the workforce costs data ([see code](https://github.com/Public-Health-Scotland/NRAC/commit/95408e57bcf23dbe1aaf479e1cd8e8972dccfaf0)).

```{r}

add_age_labels <- function(.){
  . |> 
  mutate(agegrp_label = case_when(
    agegrp == 1 ~ "0-1",
    agegrp == 2 ~ "2-4",
    agegrp == 3 ~ "5-9",
    agegrp == 4 ~ "10-14",
    agegrp == 5 ~ "15-19",
    agegrp == 6 ~ "20-24",
    agegrp == 7 ~ "25-29",
    agegrp == 8 ~ "30-34",
    agegrp == 9 ~ "35-39",
    agegrp == 10 ~ "40-44",
    agegrp == 11 ~ "45-49",
    agegrp == 12 ~ "50-54",
    agegrp == 13 ~ "55-59",
    agegrp == 14 ~ "60-64",
    agegrp == 15 ~ "65-69",
    agegrp == 16 ~ "70-74",
    agegrp == 17 ~ "75-79",
    agegrp == 18 ~ "80-84",
    agegrp == 19 ~ "85-89",
    agegrp == 20 ~ "90+",
    TRUE ~ NA
                )
         )
}

add_sex_label <- function(.){
  . |> mutate(sex_label = ifelse(sex == 1, "Male", "Female"))
}

add_staff_label <- function(.){
  . |> 
    mutate(staff_type_label = case_when(
      staff_type == "ahp_staff_direct_supplies_number" ~ "AHP", 
      staff_type == "medical_dental_staff_number" ~ "Medical & Dental", 
      staff_type == "nursing_staff_number" ~ "Nursing", 
      staff_type == "other_direct_care_number" ~ "Other Direct Care", 
      staff_type == "pharmacy_staff_direct_supplies_number" ~ "Pharmacy", 
      staff_type == "theatre_staff_direct_supplies_number" ~ "Theatre",
      TRUE ~ NA
    )
  )
}
  
workforce_cost_per_head <- readRDS("data/acute_cost_per_head_workforce_v26_27.rds") |> 
  mutate(acute_cph = round(acute_cph, 1)) |> 
  add_staff_label() |> 
  add_age_labels() |> 
  add_sex_label()


workforce_cost_per_head |> select(staff_type_label, agegrp_label, sex_label, acute_cph)
```

## Acute Costs Curves by Staff Type

Now that we have the cost per head data by staff type, age-group & sex, let's plot the costs curves.

```{r}
library(rlist)
library(ggplot2)
library(patchwork)
library(ggiraph)
library(viridis)

staff_types <- unique(workforce_cost_per_head$staff_type_label)

x_axis_labels <- workforce_cost_per_head |> 
  arrange(agegrp) |> 
  select(agegrp, agegrp_label) |> 
  distinct() |> 
  pull(agegrp_label)

# create a cost-curves plot for each staff type
acute_plots <- list()

for(i in 1:length(staff_types)){
  
  staff_type_i <- staff_types[i]

  p <- ggplot(workforce_cost_per_head |> 
                filter(staff_type_label == staff_type_i),
              aes(x = agegrp, y = acute_cph, color = sex_label, group = sex_label, data_id = sex_label))+
    
    # plot points and lines
    geom_point_interactive(aes(
      shape = sex_label,
      tooltip = paste0("Sex: ", sex_label, "\nAge: ", agegrp_label, "\nCPH: ", acute_cph)))+
    # geom_line_interactive(aes(group = sex_label))+
    geom_line_interactive()+
    
    
    # labels, colors, themes
    scale_x_continuous(breaks = c(1:length(x_axis_labels)), labels = x_axis_labels)+
    labs(title = staff_type_i, x = "Age Group", y = "Cost per Head")+
    scale_color_manual_interactive(values = viridis(begin = 0.25, end = 0.75, n = 2))+
    theme_minimal()+
    theme(axis.text.x = element_text(size = 6, angle = 45), 
          plot.title = element_text(size = 10)) 
  
  # add plot to list  
  acute_plots <- list.append(acute_plots, p)
  
  names(acute_plots)[i] <- staff_type_i
}

# combine all the plots
patchwork <- wrap_plots(acute_plots, 
                        ncol = 2,
                        nrow = 3,
                        axis_titles = "collect", 
                        guides = "collect")


# CSS options - separate hover and selection styles
hover_css <- "
  filter: brightness(75%);
  cursor: pointer;
  transition: all 0.5s ease-out;
  stroke-width: 1.5px;
  fill-opacity: 0.8;
"

selection_css <- "
  filter: brightness(1.15);
  stroke-width: 2px;
  stroke: #FF5722;
  fill-opacity: 1;
"

inv_css <- "opacity:0.5; transition: all 0.2s ease-out;"

# Add interactive tooltips with proper selection
girafe(ggobj = patchwork, options = list(
  opts_hover(css = hover_css),
  opts_tooltip(css = "background-color:lightgray; color:black; border-radius:10px;"),
  opts_selection(
    css = selection_css,
    type = "single",  # or "line" if you want to select lines
    linked = TRUE
  ),
  opts_selection_inv(css = inv_css)
))
```

## Next steps

-   load the output explore and make charts

-   run a version with total workforce costs

## Questions on methodology

-   Fixed vs Variable split? The [NRAC acute costs weight script](https://github.com/Public-Health-Scotland/NRAC/blob/develop/R/2-HCHS/1-age-sex/2-cost-weights/1-acute_cost_weights.R) uses fixed and variable cost proportions to derive a total cost based on whether the patient is an inpatient or daycase. However these proportions were derived from a SPSS syntax using PLICS data which does not seem to include workforce costs. This raises the question on whether a these fixed and variable cost ratios can be applied to the workforce costs in SFR 5.3. Ask SG if they how they plan to use the workforce cost per head, do they plan to use it as proportion of total costs?

-   Cost per specialty, or cost for all specialties?

## Caveats Summary

-   Workforce costs are assumed to be 100% fixed costs.
-   NHS Borders costs are excluded from these data since they were unable to submit a full SFR 5.3 dataset for financial year 23/24 (see [23/24 Scottish Health Service Costs](https://publichealthscotland.scot/media/34037/2025-02-25-costs-report_final.pdf) p.21).
-   Combines workforce costs from SFRs: 5.3, 5.5, 5.7N, 5.7.
-   Use all the activity data A&E, SMR00, and SMR01.

