---
title: "Workforce Cost Curves: Acute care programme"
format: 
   html:
     df-print: paged
     toc: true
---

**Main Goal: Plot the cost per head by age band and sex, for the NRAC acute care programme. These outputs rely on the NRAC cost per head calculations**

Throughout this report we split this main goal into smaller goals to make things easier to digest.

## Set-up

```{r}
#| output: false

# set-up

# data retrieval and wrangling tools
library(rlist)
library(readr)
library(janitor)
library(dplyr)
library(tidyr)
library(odbc)

# data viz tools for pretty interactive graphs
library(ggplot2)
library(patchwork)
library(ggiraph)
library(viridis)

# load custom functions
for (filename in list.files(
  "functions", pattern = "\\.R", full.names = TRUE, 
  recursive = TRUE)
) {
  source(filename)
}

# load reference files
cbdcs_column_dictionary_sfr_53 <- read_csv("reference_files/cbdcs_column_dictionary.csv", show_col_types = FALSE) |> 
  clean_names() |> 
  filter(cd_formname == "SFR 5.3") |> 
  remove_constant() |> 
  mutate(cd_snakecase_description = make_clean_names(cd_description), 
         cd_columnno = as.character(paste0("col", cd_columnno)))

#if outputs directory doesnt exist create it
if(!(dir.exists("./outputs"))){
  dir.create("./outputs")
}

```

## Reproduce the NRAC 26/27 Acute costs curve for all Acute costs

Load the acute cost per head output from the NRAC 26/27 publication. Then, plot the cost per head by age group and sex.

The acute cost per head output is sourced from <https://github.com/Public-Health-Scotland/NRAC/blob/develop/R/2-HCHS/1-age-sex/2-cost-weights/1-acute_cost_weights.R>.

```{r}

total_exp_cost_per_head <- readRDS("data/acute_cost_per_head_v26_27.rds") |>
  mutate(acute_cph = round(acute_cph, 2)) |> 
  add_age_labels() |> 
  add_sex_label()

p <- plot_cost_curves(total_exp_cost_per_head, x_title = "Acute cost per head by age and sex for total expenditure")

girafe_plot_options(p)
```

Let's attempt to produce a similar output for workforce costs only.

## 23/24 Acute Workforce Costs Data

Goal: The goal in this section is to calculate total workforce costs by line number, following a similar methodology as in the [NRAC hospital activity costs](https://github.com/Public-Health-Scotland/NRAC/blob/develop/R/2-HCHS/1-age-sex/1-costs/3-hospital_activity_costs.R)

We source the necessary workforce costs data from various Scottish Financial Returns (SFR), see table below:

| Form Name | Description | Columns |
|-------------------|-------------------|----------------------------------|
| SFR 5.3 | Inpatient costs | col 3 (number of discharges), col 7-12 (costs by staff type) |
| SFR 5.5 | Day Case costs | col 1 (number of attendances), col 5-10 (costs by staff type) |
| SFR 5.7 | Outpatients treated in consultant clinics | col 1 (number of attendances), col 6-11 (costs by staff type) |
| SFR 5.7N | Outpatients treated at nurse-led clinics | col 1 (number of attendances), col 4-7 (costs by staff type) |
| SFR 5.9 | Day Patients | col 1 (number of attendances), col 4-8 (costs by staff type) |

***NHS Borders costs are excluded from these data since they were unable to submit a full SFR dataset for financial year 23/24 (see [23/24 Scottish Health Service Costs](https://publichealthscotland.scot/media/34037/2025-02-25-costs-report_final.pdf) p.21)***

```{r}
id_df <- read_csv(
  "/PHI_conf/Finance/CostsBook/Reference Files/userid.csv",
  col_types = "c"
)
uid <- id_df$username
pwd <- id_df$password

id_colnames <- c("formname", "provider", "location", "lineno")

# get costs data
sfr_53 <- get_costs_data(2023, "SFR 5.3") |>
  select(all_of(id_colnames), col3, c(col7:col12)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

# rename columns to something humans can understand
colnames(sfr_53) <- human_cbdcs_colnames(sfr_53) 

sfr_55 <- get_costs_data(2023, "SFR 5.5")|> 
  select(all_of(id_colnames), col1, c(col5:col10)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_55) <- human_cbdcs_colnames(sfr_55)
# sfr_55

sfr_57 <- get_costs_data(2023, "SFR 5.7")|> 
  select(all_of(id_colnames), col1, c(col6:col11)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_57) <- human_cbdcs_colnames(sfr_57)
# sfr_57

sfr_57n <- get_costs_data(2023, "SFR 5.7N")|> 
  select(all_of(id_colnames), col1, c(col4:col7)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_57n) <- human_cbdcs_colnames(sfr_57n)
# sfr_57n

sfr_59 <- get_costs_data(2023, "SFR 5.9")|> 
  select(all_of(id_colnames), col1, c(col4:col8)) |>
  # remove rows where all pay columns are have 0 costs
  filter(!if_all(where(is.numeric), ~ .x==0))

colnames(sfr_59) <- human_cbdcs_colnames(sfr_59)
# sfr_59


```

Here's an example of what the SFR data looks like.

```{r}
sfr_53
```

Next, let's recode SFR line numbers to map these costs to the correct specialties in the Scottish Morbidity Records(SMR) activity data. Finally, let's aggregate the workforce costs at specialty level (ie. aggregate costs by line number and patient type). Here we assume that the proportion of the costs that are fixed is 1. We assume this because we have limited data on which proportion of the workforce costs reported are fixed (eg. core staff), and which are variable (eg. NHS staff bank, overtime, etc.). This is all neatly packaged in the function `parse_cost_by_lineno` and iterated over all the SFRs.

```{r}
sfr_list <- list(sfr_53, sfr_55, sfr_57, sfr_57n, sfr_59)

output_list <- lapply(sfr_list, parse_cost_by_lineno)

workforce_costs_by_lineno <- do.call(rbind, output_list)
workforce_costs_by_lineno

saveRDS(workforce_costs_by_lineno, "outputs/workforce_costs_by_lineno.rds")
```

## Calculate cost per head

Now that the data is aggregated by patient type and line number, we can more accurately match the costs for the financial year to A&E, Outpatients (SMR00) and Inpatients & Day Cases (SMR01) activity data. The activity data is used to derive a total cost per age-sex group, then these costs are divided by the mid year population estimates, giving an estimated cost per head.

The above is executed by a modified version of the NRAC Acute costs weights script which I adapted to read-in the workforce costs data ([see code](https://github.com/Public-Health-Scotland/NRAC/commit/95408e57bcf23dbe1aaf479e1cd8e8972dccfaf0)).

```{r}

workforce_cost_per_head <- readRDS("data/acute_cost_per_head_workforce_all_staff_v26_27.rds") |> 
  mutate(acute_cph = round(acute_cph, 1)) |> 
  add_staff_label() |> 
  add_age_labels() |> 
  add_sex_label()


workforce_cost_per_head |> select(staff_type_label, agegrp_label, sex_label, acute_cph)
```

## Acute Costs Curves by Staff Type

Now that we have the cost per head data by staff type, age-group & sex, let's plot the costs curves. The code below loops through each staff type and line chart showing the cost per head by age group, split between males and females.

```{r}

staff_types <- unique(workforce_cost_per_head$staff_type_label)

# create a cost-curves plot for each staff type
acute_plots <- list()

for(i in 1:length(staff_types)){
  
  staff_type_i <- staff_types[i]
  
  data_i <- workforce_cost_per_head |> 
                filter(staff_type_label == staff_type_i)
  
  plot_i <- plot_cost_curves(x = data_i, x_title = staff_type_i)
  
  # add plot to list  
  acute_plots <- list.append(acute_plots, plot_i)
  
  names(acute_plots)[i] <- staff_type_i
}

```

Let's take a look at the costs curves for Acute across all staff types.

```{r}

girafe_plot_options(acute_plots[[which(names(acute_plots) == "All Staff")]])
```

Breakdown of the cost per head by staff type.

```{r}
# combine all the plots except for all staff
patchwork <- wrap_plots(acute_plots[which(names(acute_plots) != "All Staff")], 
                        ncol = 2,
                        nrow = 3,
                        axis_titles = "collect", 
                        guides = "collect")

girafe_plot_options(patchwork)

```

## Questions on methodology

-   What are these costs curves going to be used for?

    -   SR overlayed the acute costs curves with population curves, if we isolate solely workforce costs does that change the picture? Looking to predict workforce needs based on population/demand.
    -   JP is trying to estimate the burden of disease and workforce demands, looking to do some modelling on how burden of disease will impact workforce demand.

-   Fixed vs Variable split? The [NRAC acute costs weight script](https://github.com/Public-Health-Scotland/NRAC/blob/develop/R/2-HCHS/1-age-sex/2-cost-weights/1-acute_cost_weights.R) uses fixed and variable cost proportions to derive a total cost based on whether the patient is an inpatient or daycase. However these proportions were derived from a SPSS syntax using PLICS data which does not seem to include workforce costs. This raises the question on whether a these fixed and variable cost ratios can be applied to the workforce costs in SFR 5.3. Ask SG if they how they plan to use the workforce cost per head, do they plan to use it as proportion of total costs?

    -   if not too time consuming try re running the analysis using the existsing fixed/variable split

-   Include Laboratory recharge costs? According to the costsbook user manual they include all costs such as: supplies, staff, and allocated costs such as heat, light and capital charges.

    -   No labs costs needed.

## Caveats Summary

-   Workforce costs are assumed to be 100% fixed costs.
-   NHS Borders costs are excluded from these data since they were unable to submit a full SFR 5.3 dataset for financial year 23/24 (see [23/24 Scottish Health Service Costs](https://publichealthscotland.scot/media/34037/2025-02-25-costs-report_final.pdf) p.21).
-   Combines workforce costs from SFRs: 5.3, 5.5, 5.7N, 5.7.
-   Uses activity data A&E, SMR00, and SMR01 to derive total costs.
-   Apart from Medical, Dental and Nursing, all other staff costs include supplies.
-   Cost per head is derived by dividing the total costs by mid year population estimates from NRS.

## Next Steps

-   Overlay workforce costs curves with acute expenditure.

-   Re-run with the exists fixed/variable costs breakdown used in NRAC.

-   Meet with SR week starting 29th March, and send JP a `.csv` of the acute workforce cost per head in relational format.

-   TBC: get costs-curve for the care of the elderly program for SR.

-   Bonus: publish and share Quarto report with a password.
